---
title: "Getting started with IQCC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with IQCC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(IQCC)
```

## What IQCC is for

IQCC provides improved quality-control charts for settings in which familiar
normal, three-sigma limits can be poorly calibrated. These settings include
small subgroups, rare nonconformities, discrete or skewed statistics, and
multivariate variability. The package emphasizes exact limits where they are
available, distributional corrections where they are justified, and explicit
calculation of the actual false-alarm risk.

IQCC complements `qcc`. The latter provides a broad statistical process-control
framework, and IQCC reuses it for several plots. IQCC adds specialized
numerical functions and plotting wrappers for corrected attribute charts,
double-sampling plans, exact range limits, generalized variance, and the
auxiliary trace statistic. Use `qcc` for a general SPC workflow; use IQCC when
the calibration of a particular small-sample or non-normal statistic matters.

## Numerical functions and chart wrappers

The preferred workflow is to calculate and inspect a design before drawing a
chart. For example, `pchart_limits()` returns the limits without plotting, and
`pchart_alpha_risk()` evaluates their exact binomial false-alarm probability.
The wrapper `cchart.p()` applies the selected method to observed subgroups and
returns the underlying `qcc` object invisibly.

```{r numerical-before-plotting}
p_limits <- pchart_limits(p = 0.015, n = 20, type = "cf2")
p_risk <- pchart_alpha_risk(
  p = 0.015,
  n = 20,
  lcl = p_limits$lcl,
  ucl = p_limits$ucl
)

data.frame(
  lcl = p_limits$lcl,
  center = p_limits$center,
  ucl = p_limits$ucl,
  actual_alpha = p_risk
)
```

The same separation is available for p and u limits, DS-np operating
characteristics, generalized variance, and `tr(V)`. The older R, S, and
Hotelling interfaces remain primarily chart-oriented; their plotting functions
should not be mistaken for generic numerical limit APIs.

## Phase I, Phase II, and reference parameters

Phase I data establish or assess the in-control reference. Phase II data are
new subgroups monitored against limits fixed from that reference. IQCC uses
method-specific argument names:

| Chart | Phase I | Phase II | Known reference |
|:--|:--|:--|:--|
| p | `x1` or `p1` | `x2` or `p2` | `phat` |
| u | `x1` or `u1` | `x2` or `u2` | `lambda` |
| R | `y` for exact limits | `x` | none in the wrapper |
| Hotelling T-squared | `data.1()`, `stats()`, `T2.1()` | `data.2()`, `T2.2()` | Phase I estimates |
| generalized variance | `x1` | `x2` | `Sigma` |
| `tr(V)` | `x` | `newdata` | `Sigma0` |

When `phat`, `lambda`, `Sigma`, or `Sigma0` is supplied, it is treated as an
external in-control reference. Otherwise, the relevant wrapper estimates it
from Phase I. Plug-in limits based on estimated parameters do not automatically
account for the additional estimation uncertainty.

## Input formats

Attribute charts accept one-dimensional vectors. Counts are paired with their
sample or inspection sizes; proportions and rates can be supplied directly.

```{r attribute-inputs}
p_counts <- c(0, 1, 0, 2, 1, 0)
p_sizes <- c(40, 50, 45, 60, 55, 50)
p_proportions <- p_counts / p_sizes

u_counts <- c(1, 0, 3, 2, 1, 4)
u_sizes <- c(10, 12, 15, 11, 14, 13)
u_rates <- u_counts / u_sizes

data.frame(
  p_count = p_counts,
  p_proportion = p_proportions,
  u_count = u_counts,
  u_rate = u_rates
)
```

Multivariate variability functions accept a list of subgroup matrices, a
subgroup-by-observation-by-variable array, or a matrix made by stacking equal
subgroups. Here all three representations contain the same data.

```{r multivariate-inputs}
g1 <- cbind(
  x = c(-1, 0, 1, -1, 1),
  y = c(-1, 1, 0, 1, -1)
)
g2 <- sweep(g1, 2, c(0.2, -0.1), "+")
g3 <- g1 %*% matrix(c(1.1, 0.1, 0, 0.9), nrow = 2)
groups <- list(g1, g2, g3)

group_array <- array(NA_real_, dim = c(3, 5, 2))
for(i in seq_along(groups))
  group_array[i, , ] <- groups[[i]]

stacked_groups <- do.call(rbind, groups)

cbind(
  from_list = gv_stat(groups),
  from_array = gv_stat(group_array),
  from_stacked_matrix = gv_stat(stacked_groups, size = 5)
)
```

The legacy Hotelling helpers use a different array convention:
observation-by-variable-by-subgroup. Check `dim()` before passing an array
between the two families.

## Corrected p and u charts

The numerical p and u functions support normal, first-order Cornish-Fisher
(`"cf1"`), and second-order Cornish-Fisher (`"cf2"`) limits. Their chart
wrappers additionally support standardized observations.

```{r corrected-attribute-limits}
p_methods <- do.call(
  rbind,
  lapply(c("normal", "cf1", "cf2"), function(method) {
    lim <- pchart_limits(0.02, n = 40, type = method)
    data.frame(method = method, lcl = lim$lcl, ucl = lim$ucl,
               applicable = lim$applicable)
  })
)

u_methods <- do.call(
  rbind,
  lapply(c("normal", "cf1", "cf2"), function(method) {
    lim <- uchart_limits(0.20, n = 10, type = method)
    data.frame(method = method, lcl = lim$lcl, ucl = lim$ucl)
  })
)

p_methods
u_methods
```

The p-chart workflow below estimates the in-control proportion from Phase I,
computes the Phase II limits explicitly, and then constructs the Phase II
chart. This makes the sequence from reference data to monitoring data visible.

```{r p-chart, fig.show='hold'}
p_phase1_counts <- c(1, 0, 1, 2, 0, 1, 1, 0)
p_phase1_sizes <- rep(50, length(p_phase1_counts))
p_phase1_estimate <- sum(p_phase1_counts) / sum(p_phase1_sizes)
p_phase2_limits <- pchart_limits(
  p = p_phase1_estimate,
  n = p_sizes,
  type = "cf2"
)

p_chart <- cchart.p(
  type = "cf2",
  x1 = p_phase1_counts,
  n1 = p_phase1_sizes,
  x2 = p_counts,
  n2 = p_sizes
)
c(
  estimated_phat = p_phase1_estimate,
  smallest_phase2_ucl = min(p_phase2_limits$ucl),
  largest_phase2_ucl = max(p_phase2_limits$ucl)
)
```

The u chart uses a known rate and direct rate input to demonstrate the
prospective alternative.

```{r u-chart, fig.show='hold'}
u_chart <- cchart.u(
  type = "standardized",
  u2 = u_rates,
  n2 = u_sizes,
  lambda = 0.15
)
class(u_chart)
```

For estimated parameters, supply Phase I counts or proportions/rates instead.
With unequal sizes, IQCC uses pooled estimators: `sum(counts) / sum(sizes)`.

## Double-sampling np

A DS-np plan can accept or signal after the first sample, or inspect a second
sample when the first count falls in the warning region. The pure functions
calculate acceptance probability, ARL, and ASS for explicit limits.

```{r dsnp-performance}
ds_plan <- list(
  p0 = 0.005,
  p1 = 0.010,
  n1 = 50,
  n2 = 242,
  wl = 1.5,
  ucl1 = 2.5,
  ucl2 = 4.5
)

data.frame(
  arl0 = with(ds_plan, dsnp_arl(p0, n1, n2, wl, ucl1, ucl2)$arl),
  arl1 = with(ds_plan, dsnp_arl(p1, n1, n2, wl, ucl1, ucl2)$arl),
  ass0 = with(ds_plan, dsnp_ass(p0, n1, n2, wl, ucl1)$ass)
)
```

The chart object records every stage decision as well as its design and
operating characteristics.

```{r dsnp-chart}
ds_chart <- cchart.DSnp(
  x1 = c(0, 1, 2, 3, 1),
  x2 = c(NA, NA, 2, NA, NA),
  n1 = 10,
  n2 = 20,
  p0 = 0.05,
  p1 = 0.10,
  wl = 1.5,
  ucl1 = 2.5,
  ucl2 = 4.5,
  plot = FALSE
)
ds_chart$data
ds_chart$performance
```

`dsnp_limits()` searches limits for fixed `n1` and `n2`; `dsnp_design()`
searches across supplied sample-size ranges. These are exhaustive discrete
searches over the requested ranges, not continuous optimization routines.

## Range and standard-deviation charts

Rows are subgroups and columns are observations for the R and S wrappers. The
range chart offers conventional (`"norm"`) and exact relative-range
(`"tukey"`) limits. The S chart offers normalized (`"n"`) and exact
chi-square (`"e"`) limits.

```{r range-chart}
data(pistonrings)
r_chart <- cchart.R(
  x = pistonrings[26:32, ],
  n = 5,
  type = "tukey",
  y = pistonrings[1:25, ]
)
class(r_chart)
```

```{r s-chart}
data(softdrink)
s_chart <- cchart.S(softdrink, type = "e", m = 10)
class(s_chart)
```

Use `alpha.risk(n)` to inspect the actual risk of the conventional R chart,
and `table.qtukey()` to inspect relative-range quantiles. The current R and S
wrappers do not expose a complete pure numerical limits object analogous to
`pchart_limits()`.

## Hotelling T-squared

The Hotelling workflow estimates the Phase I mean and covariance, computes a
T-squared statistic for each subgroup, and evaluates later subgroups against
the Phase II design. Its simulation helpers return arrays in
observation-by-variable-by-subgroup order.

```{r hotelling, fig.show='hold'}
set.seed(2026)
hotelling_phase1 <- data.1(
  m = 8,
  n = 5,
  mu = c(0, 0),
  Sigma = diag(2)
)
hotelling_reference <- IQCC::stats(hotelling_phase1, m = 8, n = 5, p = 2)
hotelling_t2_phase1 <- T2.1(hotelling_reference, m = 8, n = 5)

hotelling_phase2 <- data.2(hotelling_reference, n = 5, p = 2)
hotelling_t2_phase2 <- T2.2(
  hotelling_phase2,
  hotelling_reference,
  n = 5
)

cchart.T2.1(hotelling_t2_phase1, m = 8, n = 5, p = 2)
c(phase1_max = max(hotelling_t2_phase1), phase2 = hotelling_t2_phase2)
```

`cchart.T2.2()` draws the operational Phase II chart. The Phase I and Phase II
limits differ, so the two plotting functions are not interchangeable.

## Generalized variance and tr(V)

For each multivariate subgroup, `gv_stat()` computes the determinant of the
sample covariance matrix. `trv_stat()` computes
`(n - 1) tr(Sigma0^-1 S)`. The determinant and standardized trace detect
different covariance changes and are complementary.

```{r multivariate-statistics}
data.frame(
  subgroup = seq_along(groups),
  generalized_variance = gv_stat(groups),
  trace_statistic = trv_stat(groups, Sigma0 = diag(2))
)
```

Dimension-two generalized-variance limits have an exact implementation.
Normal, Cornish-Fisher, and simulation methods are also available. Simulation
is reproducible when `seed` is supplied.

```{r generalized-variance-methods}
gv_method_table <- do.call(
  rbind,
  lapply(c("normal", "cf", "exact"), function(method) {
    lim <- gv_limits(n = 5, p = 2, det_sigma = 1, type = method)
    data.frame(method = method, center = lim$center, ucl = lim$ucl)
  })
)
gv_simulated <- gv_limits(
  n = 5,
  p = 2,
  det_sigma = 1,
  type = "simulation",
  nsim = 2000,
  seed = 2026
)

gv_method_table
gv_simulated[c("type", "ucl", "nsim", "seed")]
```

The trace statistic has an exact chi-square upper limit under a known
in-control covariance matrix. Its simulation option is mainly a diagnostic
comparison with that same null distribution.

```{r multivariate-charts}
gv_chart <- cchart.GV(
  x1 = groups[1:2],
  x2 = groups[3],
  type = "cf",
  plot = FALSE
)

trv_chart <- cchart.trV(
  x = groups[1:2],
  newdata = groups[3],
  Sigma0 = diag(2),
  type = "chisq",
  plot = FALSE
)

summary(gv_chart)
summary(trv_chart)
```

If `Sigma` or `Sigma0` is omitted, Phase I covariance information is used as a
plug-in estimate. Phase II observations never update that reference. The
nominal known-parameter distribution does not include the extra uncertainty
introduced by estimating the reference from a finite Phase I sample.

## Choosing a method

- `normal` uses conventional or moment-matched Gaussian limits.
- `cf1` and `cf2` are first- and second-order Cornish-Fisher methods for p and
  u charts. Generalized variance uses `type = "cf"` with `cf_order = 1` or `2`.
- `exact` refers to a supported distributional result, not to simulation. Its
  scope depends on the statistic and dimension.
- `standardized` is available in the p and u plotting wrappers and places
  subgroup observations on a z scale.
- `simulation` uses Monte Carlo quantiles. Set `nsim` and `seed`, and report
  simulation error when using the result scientifically.
- The exact `tr(V)` method is named `"chisq"`, reflecting its chi-square null
  distribution.

An exact known-parameter limit can become a plug-in operational limit when its
reference parameter is estimated. Report that distinction explicitly.

## Inspecting limits and signals

Modern IQCC chart objects are lists. Inspect their components rather than
reading values from a plot.

```{r object-inspection}
data.frame(
  chart = c("DS-np", "generalized variance", "tr(V)"),
  n_statistics = c(
    nrow(ds_chart$data),
    length(gv_chart$statistics),
    length(trv_chart$statistics)
  ),
  n_signals = c(
    sum(ds_chart$data$signal),
    length(gv_chart$out.of.control),
    length(trv_chart$out.of.control)
  )
)

gv_chart$limits[c("lcl", "center", "ucl", "type", "alpha")]
```

For p and u charts, inspect the returned `qcc` object, including its
`statistics`, `limits`, and `violations` components. For DS-np, `|S|`, and
`tr(V)`, `summary()` provides a compact account of parameters, phases, limits,
and signals.

IQCC generally signals strict excursions beyond a limit. For generalized
variance, a point signals when it is below the LCL or above the UCL; equality
does not signal. The `tr(V)` chart signals only when the statistic is strictly
above its UCL. DS-np uses integer thresholds derived from its fractional
limits, so inspect `ds_chart$limits` before interpreting a boundary count.

## Common errors

- Do not supply both counts and proportions/rates for the same attribute-chart
  phase.
- Match every count vector with a sample-size vector of the same length, unless
  a scalar size is intentionally recycled.
- Do not use `p1` to mean the same thing everywhere: in `cchart.p()` it is a
  vector of Phase I proportions, while in DS-np functions it is the scalar
  out-of-control process proportion.
- For DS-np, provide a second-stage count whenever the first-stage count lies
  in the warning zone; use `NA` only when the second sample is not required.
- For multivariate variability, use subgroup-by-observation-by-variable arrays;
  the legacy Hotelling helpers use observation-by-variable-by-subgroup arrays.
- Generalized variance requires more observations than variables in every
  subgroup. `Sigma` and `Sigma0` must be symmetric positive-definite matrices
  with the correct dimension.
- Do not describe a simulated limit as exact, and do not assume a plug-in limit
  has known-parameter coverage.

## Further documentation

Read `vignette("high-quality-processes", package = "IQCC")` for corrected p
charts and DS-np plans, `vignette("statistical-foundations", package =
"IQCC")` for formulas and published validations, and
`vignette("iqcc-positioning", package = "IQCC")` for the package's scope.
Function manuals contain the complete input validation, return values,
decision inequalities, and method-specific limitations.
