---
title: "High-quality processes and rare nonconformities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{High-quality processes and rare nonconformities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(IQCC)
```

## Motivation

High-quality processes are processes in which nonconformities are rare. In that
setting, attribute data are highly discrete, bounded below by zero, and often
strongly asymmetric. Classical normal-approximation control limits may therefore
be poorly calibrated, even when the usual chart is familiar and easy to apply.

IQCC addresses this problem with corrected p-chart limits, exact binomial
false-alarm calculations, and a complete double-sampling np workflow including
performance evaluation, automatic limit search, chart construction, and plotting.

## Cornish-Fisher corrected p charts

IQCC distinguishes two corrected methods:

- `type = "cf1"` uses the first skewness correction;
- `type = "cf2"` uses the two-adjustment operational limits documented by
  Joekes and Barbosa (2013).

The following chunk is executed and compares all three limit methods.

```{r p-chart-limits}
methods <- c("normal", "cf1", "cf2")
p_results <- do.call(
  rbind,
  lapply(methods, function(method) {
    lim <- pchart_limits(p = 0.015, n = 20, type = method)
    risk <- pchart_alpha_risk(
      p = 0.015,
      n = 20,
      lcl = lim$lcl,
      ucl = lim$ucl
    )
    data.frame(
      method = method,
      lcl = lim$lcl,
      center = lim$center,
      ucl = lim$ucl,
      actual_alpha = risk$alpha,
      arl0 = risk$arl0
    )
  })
)
p_results
```

```{r p-chart-figure}
data(binomdata)
cchart.p(
  x1 = binomdata$Di[1:12],
  n1 = binomdata$ni[1:12],
  type = "cf2",
  x2 = binomdata$Di[13:25],
  n2 = binomdata$ni[13:25]
)
```

## Double-sampling np charts

A double-sampling np chart uses two possible sampling stages. A first sample is
inspected. If the count of nonconforming items is clearly acceptable, the process
is not signaled. If it is clearly large, the chart signals immediately. If the
count lies in an intermediate region, a second sample is inspected and the
combined count is used for the final decision.

IQCC provides the complete DS-np workflow:

- `dsnp_prob_accept()` computes the total acceptance probability;
- `dsnp_arl()` computes average run length;
- `dsnp_ass()` computes average sample size;
- `dsnp_limits()` searches and ranks feasible limits;
- `cchart.DSnp()` classifies observations and constructs the chart.

## Inspecting a published plan

```{r dsnp-core}
n1 <- 34
n2 <- 162
wl <- 1.5
ucl1 <- 2.5
ucl2 <- 4.5
p0 <- 0.005
p1 <- 0.0075

published_plan <- data.frame(
  metric = c("P(accept | p0)", "ARL0", "ARL1", "ASS0"),
  value = c(
    dsnp_prob_accept(p0, n1, n2, wl, ucl1, ucl2)$pt,
    dsnp_arl(p0, n1, n2, wl, ucl1, ucl2)$arl,
    dsnp_arl(p1, n1, n2, wl, ucl1, ucl2)$arl,
    dsnp_ass(p0, n1, n2, wl, ucl1)$ass
  )
)
published_plan
```

The approximate published performance values are `ARL0 = 803.41`,
`ARL1 = 193.22`, and `ASS0 = 35.94`.

## Searching for limits

A reduced search space is used here to exercise the full algorithm without
making documentation builds unnecessarily slow.

```{r dsnp-search}
lim <- dsnp_limits(
  p0 = 0.05,
  n1 = 5,
  n2 = 10,
  alpha = 0.05,
  p1 = 0.10,
  max_results = 5
)
lim$best[, c("wl", "ucl1", "ucl2", "p_signal0",
             "arl0", "arl1", "ass0")]
```

## Interpreting fractional limits

For a DS-np plan with first-stage count `D1` and second-stage count `D2`, IQCC
uses the following rule:

- accept at the first stage when `D1 <= floor(wl)`;
- signal at the first stage when `D1 >= floor(ucl1) + 1`;
- otherwise inspect the second sample;
- after the second sample, accept when `D1 + D2 <= floor(ucl2)`;
- signal after the second stage otherwise.

## Constructing the chart

```{r dsnp-chart}
x1 <- c(0, 1, 2, 3, 1, 0, 2, 4, 1, 0)
x2 <- c(NA, NA, 2, NA, NA, NA, 3, NA, NA, NA)

chart <- cchart.DSnp(
  x1,
  n1 = 10,
  n2 = 20,
  p0 = 0.05,
  x2 = x2,
  wl = 1.5,
  ucl1 = 2.5,
  ucl2 = 4.5,
  p1 = 0.10,
  plot = TRUE
)

chart$limits
chart$performance
chart$data
```

```{r execution-marker, results='asis'}
cat("<!-- IQCC_EXECUTED_HIGH_QUALITY -->\n")
```

## Practical workflow

A practical high-quality-process workflow in IQCC is:

1. Compare normal, CF1, and CF2 limits with `pchart_limits()`.
2. Evaluate actual binomial risk with `pchart_alpha_risk()`.
3. Use `cchart.p()` for routine p-chart monitoring.
4. Use `dsnp_limits()` when inspection effort motivates a double-sampling plan.
5. Compare candidates using false-alarm probability, out-of-control ARL, and ASS.
6. Supply the selected design to `cchart.DSnp()` for operational monitoring.
