---
title: "High-quality processes and rare nonconformities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{High-quality processes and rare nonconformities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(IQCC)
```

## Motivation

High-quality processes are processes in which nonconformities are rare. In that
setting, attribute data are highly discrete, bounded below by zero, and often
strongly asymmetric. Classical normal-approximation control limits may therefore
be poorly calibrated, even when the usual chart is familiar and easy to apply.

IQCC addresses this problem with corrected p-chart limits, exact binomial
false-alarm calculations, and a complete double-sampling np (DS-np) workflow
including performance evaluation, automatic limit search, chart construction,
and plotting.

## When to correct the p chart

The standard p-chart three-sigma limits,

\[
UCL = \hat p + 3\sqrt{\hat p(1 - \hat p)/n},\qquad
LCL = \hat p - 3\sqrt{\hat p(1 - \hat p)/n},
\]

rely on a normal approximation to the binomial. When
\(n p(1-p)\) is small, the binomial distribution is strongly discrete and
asymmetric, and the actual false-alarm probability of the normal limits can be
far from the nominal \(0.0027\). IQCC provides two Cornish--Fisher corrections:

- `type = "cf1"` applies the first correction term, driven by binomial skewness;
- `type = "cf2"` retains CF1 and adds the next terms of the Cornish--Fisher
  expansion used by Joekes and Barbosa (2013), involving the fourth standardized
  cumulant and the squared skewness contribution.

The following table compares the nominal and actual performance for
\(p = 0.015\) and \(n = 20\):

```{r p-chart-limits}
methods <- c("normal", "cf1", "cf2")
p_results <- do.call(
  rbind,
  lapply(methods, function(method) {
    lim <- pchart_limits(p = 0.015, n = 20, type = method)
    risk <- pchart_alpha_risk(
      p = 0.015,
      n = 20,
      lcl = lim$lcl,
      ucl = lim$ucl
    )
    data.frame(
      method = method,
      lcl = lim$lcl,
      center = lim$center,
      ucl = lim$ucl,
      npq = lim$npq,
      applicable = lim$applicable,
      actual_alpha = risk,
      arl0 = ifelse(risk == 0, Inf, 1 / risk)
    )
  })
)
p_results
```

The normal limits produce an ARL0 of only about `r round(p_results$arl0[1])`
subgroups, far below the nominal 370. The CF2 correction brings the actual
risk substantially closer to the target.

```{r p-chart-figure}
data(binomdata)
cchart.p(
  x1 = binomdata$Di[1:12],
  n1 = binomdata$ni[1:12],
  type = "cf2",
  x2 = binomdata$Di[13:25],
  n2 = binomdata$ni[13:25]
)
```

### Diagnosing p-chart applicability

Joekes and Barbosa (2013) propose the following practical guidance based on
\(n p(1-p)\):

- use the normal limits when \(n p(1-p) \ge 5\);
- CF1 is supported when \(n p(1-p) \ge 0.25\);
- CF2 extends the corrected chart to \(n p(1-p) \ge 0.08\);
- below \(0.08\), another control-chart methodology is needed.

This is practical guidance rather than an exact theorem. The `applicable` field
returned by `pchart_limits()` implements these thresholds. For the example above:

```{r diagnostic}
p_results[, c("method", "npq", "applicable")]
```

When more than one method is applicable, the exact binomial false-alarm risk
should still be inspected. A DS-np plan is not selected solely because a p-chart
approximation is poor: its ARL and inspection effort must also be evaluated for
the shifts that matter operationally.

## Double-sampling np charts

A double-sampling np chart uses two possible sampling stages. A first sample of
size \(n_1\) is inspected. If the count of nonconforming items is at most
\(\lfloor WL \rfloor\), the process is accepted. If it is at least
\(\lfloor UCL_1 \rfloor + 1\), the chart signals immediately. Otherwise the
count lies in the warning zone and a second sample of size \(n_2\) is inspected.
After the second sample, the process is accepted when
\(D_1 + D_2 \le \lfloor UCL_2 \rfloor\) and signals otherwise.

IQCC provides the complete DS-np workflow:

- `dsnp_prob_accept()` computes the total acceptance probability;
- `dsnp_arl()` computes average run length;
- `dsnp_ass()` computes average sample size, with or without curtailed
  (truncated) second-stage inspection;
- `dsnp_limits()` searches and ranks feasible fractional limits;
- `dsnp_design()` searches sample sizes and limits under explicit ARL and ASS
  constraints;
- `cchart.DSnp()` classifies observations and constructs the chart.

### Inspecting a published plan

Joekes, Smrekar and Barbosa (2015, Table 2) report a DS-np plan for
\(p_0 = 0.005\) with \(n_1 = 34\), \(n_2 = 162\),
\(WL = 1.5\), \(UCL_1 = 2.5\), and \(UCL_2 = 4.5\):

```{r dsnp-core}
n1 <- 34
n2 <- 162
wl <- 1.5
ucl1 <- 2.5
ucl2 <- 4.5
p0 <- 0.005
p1 <- 0.0075

published_plan <- data.frame(
  metric = c("P(accept | p0)", "ARL0", "ARL1", "ASS0"),
  value = c(
    dsnp_prob_accept(p0, n1, n2, wl, ucl1, ucl2)$pt,
    dsnp_arl(p0, n1, n2, wl, ucl1, ucl2)$arl,
    dsnp_arl(p1, n1, n2, wl, ucl1, ucl2)$arl,
    dsnp_ass(p0, n1, n2, wl, ucl1)$ass
  )
)
published_plan
```

The published values are ARL0 = 803.41, ARL1 = 193.22, and ASS0 = 35.94 — all
within rounding tolerance.

### Average sample size: complete versus curtailed inspection

By default, `dsnp_ass()` assumes that whenever the first-stage count enters the
warning zone, all \(n_2\) second-stage items are inspected. However, as soon as
the cumulative nonconformity count exceeds \(UCL_2\) the rejection decision is
certain, and inspection can stop. This *curtailed* (truncated) inspection
reduces the average sample size without changing the signal probability or ARL.
The `curtailed` parameter in `dsnp_ass()` enables this convention when the user
also supplies `ucl2` (the vignette example uses complete inspection; see the
function documentation for the curtailed option).

### ARL and ASS as functions of p

The performance of a DS-np plan varies with the true nonconforming proportion.
The following evaluation shows ARL and ASS across a grid of \(p\) values:

```{r dsnp-performance-curve}
p_grid <- seq(0.001, 0.03, length.out = 30)
perf <- do.call(rbind, lapply(p_grid, function(p) {
  arl <- dsnp_arl(p, n1, n2, wl, ucl1, ucl2)$arl
  ass <- dsnp_ass(p, n1, n2, wl, ucl1)$ass
  data.frame(p = p, arl = arl, ass_complete = ass)
}))
head(perf)
```

```{r dsnp-arl-curve, fig.width = 7, fig.height = 5}
plot(perf$p, pmin(perf$arl, 1000), type = "l", col = "steelblue", lwd = 2,
     xlab = "Nonconforming proportion p", ylab = "ARL (capped at 1000)",
     main = "DS-np ARL curve (n1=34, n2=162)")
abline(v = p0, lty = 2, col = "gray")
abline(v = p1, lty = 2, col = "gray")
legend("topright", legend = c("ARL", "p0", "p1"),
       col = c("steelblue", "gray", "gray"), lty = c(1, 2, 2), lwd = c(2, 1, 1))
```

The ARL drops steeply as \(p\) increases beyond \(p_0\), illustrating the
chart's ability to detect deterioration.

### Searching for limits

When no published plan is available, `dsnp_limits()` enumerates and ranks
feasible fractional limit combinations for fixed sample sizes. A compact
example:

```{r dsnp-search}
lim <- dsnp_limits(
  p0 = 0.05,
  n1 = 5,
  n2 = 10,
  alpha = 0.05,
  p1 = 0.10,
  max_results = 5
)
lim$best[, c("wl", "ucl1", "ucl2", "p_signal0",
             "arl0", "arl1", "ass0")]
```

### Searching a complete design

The full design search over \(n_1\) and \(n_2\) is performed by
`dsnp_design()`. The problem minimizes ARL1 subject to \(\text{ARL}_0 \ge
\text{arl0\_min}\) and \(\text{ASS}_0 \le \text{ass0\_max}\):

```{r dsnp-design}
design <- dsnp_design(
  p0 = 0.05,
  p1 = 0.10,
  n1_range = 5:6,
  n2_range = 8:10,
  arl0_min = 50,
  ass0_max = 6,
  objective = "arl1",
  max_results = 5
)
design$best[, c("n1", "n2", "wl", "ucl1", "ucl2",
                "ass0", "arl0", "arl1")]
```

The published tables in Joekes, Smrekar and Barbosa (2015) report selected plans
but not the complete bounds used for the searches over \(n_1\) and \(n_2\) or
every tie-breaking rule. IQCC tests recover a published plan over an explicitly
recorded local grid and validate the exhaustive algorithm independently on a
small grid.

### Trade-off between detection and sampling effort

Double sampling can reduce average inspection and improve sensitivity to small
or moderate increases in the nonconforming proportion, at the cost of a
variable workload and a more complex operating rule. In the published reference
plan above, the corresponding single-sampling chart inspects 40 units every
cycle. The DS-np chart starts with only 34 units and has complete-inspection
\(ASS_0 = 35.94\), below 40, because most decisions are made after the first
sample. When the warning zone is reached, however, a second sample of 162 units
is required, so the maximum workload in that cycle is much larger.

The `ass0_max` constraint in `dsnp_design()` lets the practitioner balance
out-of-control detection against average inspection effort. Curtailed inspection
can reduce the effective second-stage effort further without changing the signal
probability or ARL for a fixed plan.

### Interpreting fractional limits

For a DS-np plan with first-stage count \(D_1\) and second-stage count \(D_2\),
IQCC uses the following rule:

- accept at the first stage when \(D_1 \le \lfloor WL \rfloor\);
- signal at the first stage when \(D_1 \ge \lfloor UCL_1 \rfloor + 1\);
- otherwise inspect the second sample;
- after the second sample, accept when \(D_1 + D_2 \le \lfloor UCL_2 \rfloor\);
- signal after the second stage otherwise.

The fractional limits allow intermediate non-integer thresholds that are
converted to the integer decision boundaries above.

### Constructing the chart

```{r dsnp-chart}
x1 <- c(0, 1, 2, 3, 1, 0, 2, 4, 1, 0)
x2 <- c(NA, NA, 2, NA, NA, NA, 3, NA, NA, NA)

chart <- cchart.DSnp(
  x1,
  n1 = 10,
  n2 = 20,
  p0 = 0.05,
  x2 = x2,
  wl = 1.5,
  ucl1 = 2.5,
  ucl2 = 4.5,
  p1 = 0.10,
  plot = TRUE
)

chart$limits
chart$performance
chart$data
```

```{r execution-marker, results='asis'}
cat("<!-- IQCC_EXECUTED_HIGH_QUALITY -->\n")
```

## Practical workflow

A practical high-quality-process workflow in IQCC is:

1. Diagnose p-chart applicability with \(n \hat p(1-\hat p)\) and the
   `applicable` field returned by `pchart_limits()`.
2. Compare normal, CF1, and CF2 limits with `pchart_limits()`.
3. Evaluate actual binomial risk with `pchart_alpha_risk()`.
4. Use `cchart.p()` for routine p-chart monitoring when its calibration and
   sensitivity are adequate for the application.
5. Evaluate DS-np with `dsnp_limits()` or `dsnp_design()` when sensitivity to
   relevant shifts and inspection effort motivate double sampling.
6. Evaluate ARL and ASS across the relevant \(p\) range.
7. Optionally compare complete and curtailed ASS via `dsnp_ass(curtailed = TRUE)`.
8. Supply the selected design to `cchart.DSnp()` for operational monitoring.

## References

Joekes, S. and Barbosa, E. P. (2013). An improved attribute control chart for
monitoring non-conforming proportion in high quality processes. *Control
Engineering Practice*, 21, 407--412.
doi: [10.1016/j.conengprac.2012.12.005](https://doi.org/10.1016/j.conengprac.2012.12.005).

Joekes, S., Smrekar, M. and Barbosa, E. P. (2015). Extending a double sampling
control chart for non-conforming proportion in high quality processes to the
case of small samples. *Statistical Methodology*, 23, 35--49.
doi: [10.1016/j.stamet.2014.09.003](https://doi.org/10.1016/j.stamet.2014.09.003).
