---
title: "When to use IQCC: exact and corrected control charts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When to use IQCC: exact and corrected control charts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Purpose

IQCC implements *Improved Quality Control Charts*: control charts with exact,
corrected, or standardized limits for situations where classical Shewhart-type
approximations may be poorly calibrated.

The package is not intended to replace general-purpose statistical process
control packages. Instead, it is useful when the monitoring statistic is
bounded, discrete, skewed, non-normal, or based on small samples. In those
settings, the usual three-sigma limits can produce a false alarm probability
that is meaningfully different from the nominal value.

## Positioning in the R ecosystem

A practical way to position IQCC is:

> IQCC is a specialized package for exact and corrected control-chart limits,
> especially for small samples, rare defects, asymmetric statistics, and
> selected multivariate monitoring problems.

Use a general SPC package when the goal is a broad control-chart workflow. Use
IQCC when the main concern is whether the classical limits are statistically
well calibrated for the distribution and sample size at hand.

## When IQCC is especially useful

| Monitoring problem | Classical difficulty | IQCC direction |
|---|---|---|
| Range-based dispersion monitoring | Normal approximations can inflate the actual false alarm risk, especially for small subgroups. | Use exact range-chart limits based on the relative range distribution. |
| Standard deviation monitoring | The distribution of the sample standard deviation is asymmetric for small samples. | Use exact limits derived from the chi-square distribution of the sample variance. |
| Nonconforming proportion in high-quality processes | The binomial distribution is discrete, bounded, and strongly skewed when the nonconforming proportion is very small. | Use Cornish-Fisher corrected limits or DS-np methods under development. |
| Multivariate process mean monitoring | Hotelling T² requires phase-specific calibration. | Use the implemented Phase I and Phase II Hotelling T² functions. |
| Multivariate process variability | A single normal approximation for generalized variance may be poorly calibrated. | Candidate future extensions include generalized variance and auxiliary trace charts. |

## Current implemented methods

The current package includes several families of functions:

- `cchart.R()` for range charts, including exact Tukey-based limits.
- `cchart.S()` for standard deviation charts, including exact chi-square-based limits.
- `cchart.p()` for nonconforming proportion charts, including Cornish-Fisher
  corrected limits.
- `cchart.u()` for nonconformities per unit.
- `T2.1()`, `T2.2()`, `cchart.T2.1()`, and `cchart.T2.2()` for Hotelling T²
  monitoring.
- `alpha.risk()` for studying the actual false alarm risk of classical range
  charts.
- `dsnp_prob_accept()`, `dsnp_arl()`, and `dsnp_ass()` for the numerical core of
  double-sampling np charts.

## Example: using a corrected p chart

The example below uses the package's binomial data and constructs a p chart with
Cornish-Fisher corrected limits. It is intentionally short; the purpose is to
show where IQCC fits into the usual control-chart workflow.

```{r p-chart-example, eval = FALSE}
library(IQCC)

data(binomdata)

cchart.p(
  x1 = binomdata$Di[1:12],
  n1 = binomdata$ni[1:12],
  type = "CF",
  x2 = binomdata$Di[13:25],
  n2 = binomdata$ni[13:25]
)
```

## Example: inspecting the DS-np numerical core

The DS-np plotting interface is still under development, but the numerical core
can already be used to inspect the acceptance probability, average run length,
and average sample size for a proposed plan.

```{r dsnp-core-example, eval = FALSE}
library(IQCC)

plan <- list(n1 = 34, n2 = 162, wl = 1.5, ucl1 = 2.5, ucl2 = 4.5)

dsnp_prob_accept(0.005, plan$n1, plan$n2, plan$wl, plan$ucl1, plan$ucl2)
dsnp_arl(0.005, plan$n1, plan$n2, plan$wl, plan$ucl1, plan$ucl2)
dsnp_ass(0.005, plan$n1, plan$n2, plan$wl, plan$ucl1)
```

## Development principles

Future extensions should follow three principles.

First, numerical calculations should be separated from plotting functions. A
limit, probability, average run length, or false alarm risk shown in a chart
should be reproducible by a small numerical function.

Second, new methods should be validated against published tables or examples
when possible. This is particularly important for charts based on exact or
corrected distributional calculations.

Third, IQCC should remain complementary to the broader R SPC ecosystem. Its
strongest role is to provide statistically calibrated alternatives for cases
where classical approximations are simple but inaccurate.

## Related roadmap items

The main planned extensions are:

- automatic limit search and plotting for double-sampling np charts;
- generalized variance charts for multivariate process variability;
- Cornish-Fisher corrected generalized variance limits;
- auxiliary trace charts for complementary multivariate variability monitoring;
- validation fixtures reproducing numerical values from the underlying papers.

These items are tracked in the GitHub issue roadmap and should be implemented in
small, reviewable pull requests.
