---
title: "When to use IQCC: exact and corrected control charts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When to use IQCC: exact and corrected control charts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(IQCC)
```

## Purpose

IQCC implements *Improved Quality Control Charts*: control charts with exact,
corrected, standardized, or simulation-based limits for situations where
classical Shewhart-type approximations may be poorly calibrated.

The package is not intended to replace general-purpose statistical process
control packages. Instead, it is useful when the monitoring statistic is
bounded, discrete, skewed, strongly non-normal, or based on small samples. In
those settings, the usual three-sigma limits can produce a false-alarm
probability that differs meaningfully from the nominal value.

## Positioning in the R ecosystem

> IQCC is a specialized package for exact and corrected control-chart limits,
> especially for small samples, rare nonconformities, asymmetric statistics,
> and selected multivariate monitoring problems.

Use a general SPC package when the goal is a broad control-chart workflow. Use
IQCC when the main concern is whether classical limits are statistically well
calibrated for the distribution and sample size at hand.

## Current implemented methods

The current package includes:

- `cchart.R()` and `cchart.S()` for range and standard-deviation charts;
- `pchart_limits()`, `pchart_alpha_risk()`, and `cchart.p()`;
- `uchart_limits()`, `uchart_alpha_risk()`, and `cchart.u()`;
- the DS-np design and monitoring family;
- Hotelling T-squared Phase I and Phase II functions;
- generalized-variance limits, risk diagnostics, and charts.

## Executed example: corrected p-chart limits

The chunk below is evaluated when the vignette and the pkgdown site are built.
It prints the limits and their exact binomial false-alarm probability.

```{r p-chart-example}
data(binomdata)

p0 <- sum(binomdata$Di[1:12]) / sum(binomdata$ni[1:12])
lim_p <- pchart_limits(p = p0, n = 20, type = "cf2")
risk_p <- pchart_alpha_risk(
  p = p0,
  n = 20,
  lcl = lim_p$lcl,
  ucl = lim_p$ucl
)

data.frame(
  p0 = p0,
  lcl = lim_p$lcl,
  center = lim_p$center,
  ucl = lim_p$ucl,
  actual_alpha = risk_p$alpha,
  arl0 = risk_p$arl0
)
```

The operational chart itself is also executed and produces a figure in the
published article.

```{r p-chart-figure}
cchart.p(
  x1 = binomdata$Di[1:12],
  n1 = binomdata$ni[1:12],
  type = "cf2",
  x2 = binomdata$Di[13:25],
  n2 = binomdata$ni[13:25]
)
```

## Executed example: DS-np operating characteristics

The published small-sample design can be evaluated directly without an
expensive search.

```{r dsnp-core}
n1 <- 34
n2 <- 162
wl <- 1.5
ucl1 <- 2.5
ucl2 <- 4.5
p0 <- 0.005
p1 <- 0.0075

perf_dsnp <- data.frame(
  quantity = c("ARL0", "ARL1", "ASS0"),
  value = c(
    dsnp_arl(p0, n1, n2, wl, ucl1, ucl2)$arl,
    dsnp_arl(p1, n1, n2, wl, ucl1, ucl2)$arl,
    dsnp_ass(p0, n1, n2, wl, ucl1)$ass
  )
)
perf_dsnp
```

A compact search is used in the vignette so that CI remains fast while still
executing the design algorithm.

```{r dsnp-search}
small_search <- dsnp_limits(
  p0 = 0.05,
  n1 = 5,
  n2 = 10,
  alpha = 0.05,
  p1 = 0.10,
  max_results = 5
)
small_search$best[, c("wl", "ucl1", "ucl2", "arl0", "arl1", "ass0")]
```

The chart is built with explicit limits, which makes the displayed decision
rule reproducible.

```{r dsnp-figure}
x1 <- c(0, 1, 2, 3, 1, 0, 2, 4, 1, 0)
x2 <- c(NA, NA, 2, NA, NA, NA, 3, NA, NA, NA)

chart_dsnp <- cchart.DSnp(
  x1,
  n1 = 10,
  n2 = 20,
  p0 = 0.05,
  x2 = x2,
  wl = 1.5,
  ucl1 = 2.5,
  ucl2 = 4.5,
  plot = TRUE
)
chart_dsnp$performance
```

## Executed example: generalized variance

```{r gv-example}
gv_comparison <- do.call(
  rbind,
  lapply(c("normal", "cf", "exact"), function(method) {
    limits <- gv_limits(
      n = 10,
      p = 2,
      det_sigma = 0.5320,
      type = method
    )
    data.frame(method = method, lcl = limits$lcl,
               center = limits$center, ucl = limits$ucl)
  })
)
gv_comparison
```

```{r gv-figure}
set.seed(123)
gv_data <- array(rnorm(8 * 8 * 2), dim = c(8, 8, 2))
cchart.GV(
  gv_data,
  Sigma = diag(2),
  type = "exact",
  plot = TRUE
)
```

```{r execution-marker, results='asis'}
cat("<!-- IQCC_EXECUTED_POSITIONING -->\n")
```

## Development principles

Numerical calculations should remain separated from plotting functions, new
methods should be validated against published examples whenever possible, and
IQCC should remain complementary to the broader R statistical process-control
ecosystem.
