---
title: "Statistical Foundations of IQCC"
author: "Flavio Barros and contributors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Foundations of IQCC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(IQCC)
```

# Overview

IQCC focuses on control charts whose classical normal limits may be poorly
calibrated because the monitored statistic is discrete, skewed, bounded, or
strongly non-normal. The package combines three ideas:

1. compute limits on the original measurement scale;
2. evaluate actual false-alarm probabilities whenever possible;
3. validate formulas against independent calculations and published examples.

# Binomial p chart

For

\[
X\sim\operatorname{Binomial}(n,p),\qquad \hat p=X/n,
\]

\[
E(\hat p)=p,\qquad
\operatorname{Var}(\hat p)=\frac{p(1-p)}{n}.
\]

The normal limits use

\[
p\pm z\sqrt{\frac{p(1-p)}{n}}.
\]

The first Cornish-Fisher correction adds

\[
\frac{(z^2-1)(1-2p)}{6n}
\]

to the normal quantile on the original p scale.

```{r pchart-example}
pchart_table <- do.call(
  rbind,
  lapply(c("normal", "cf1", "cf2"), function(method) {
    lim <- pchart_limits(p = 0.015, n = 20, type = method)
    risk <- pchart_alpha_risk(
      p = 0.015,
      n = 20,
      lcl = lim$lcl,
      ucl = lim$ucl
    )
    data.frame(
      method = method,
      lcl = lim$lcl,
      center = lim$center,
      ucl = lim$ucl,
      alpha = risk,
      arl0 = ifelse(risk == 0, Inf, 1 / risk)
    )
  })
)
pchart_table
```

For unequal subgroup sizes, IQCC estimates the in-control proportion using the
pooled estimator

\[
\hat p_0=\frac{\sum_i X_i}{\sum_i n_i}.
\]

# Poisson u chart

Let

\[
X\sim\operatorname{Poisson}(\lambda n),\qquad U=X/n.
\]

Then

\[
E(U)=\lambda,\qquad
\operatorname{Var}(U)=\frac{\lambda}{n}.
\]

```{r uchart-example}
u_table <- do.call(
  rbind,
  lapply(c("normal", "cf1", "cf2"), function(method) {
    lim <- uchart_limits(lambda = 0.15, n = 20, type = method)
    risk <- uchart_alpha_risk(
      lambda = 0.15,
      n = 20,
      lcl = lim$lcl,
      ucl = lim$ucl
    )
    data.frame(
      method = method,
      lcl = lim$lcl,
      center = lim$center,
      ucl = lim$ucl,
      alpha = risk,
      arl0 = ifelse(risk == 0, Inf, 1 / risk)
    )
  })
)
u_table
```

# Double-sampling np chart

Let

\[
D_1\sim\operatorname{Binomial}(n_1,p),\qquad
D_2\sim\operatorname{Binomial}(n_2,p),
\]

independently. Define

\[
a=\lfloor WL\rfloor,\qquad
b=\lfloor UCL_1\rfloor+1,\qquad
c=\lfloor UCL_2\rfloor.
\]

The decision rule is:

- accept at stage 1 when \(D_1\leq a\);
- signal at stage 1 when \(D_1\geq b\);
- continue when \(a<D_1<b\);
- accept at stage 2 when \(D_1+D_2\leq c\).

Thus

\[
ARL(p)=\frac{1}{1-P_T(p)},
\qquad
ASS(p)=n_1+n_2P_2(p).
\]

```{r dsnp-example}
prob <- dsnp_prob_accept(
  p = 0.005,
  n1 = 34,
  n2 = 162,
  wl = 1.5,
  ucl1 = 2.5,
  ucl2 = 4.5
)

data.frame(
  pt = prob$pt,
  p_signal = prob$p_signal,
  arl0 = dsnp_arl(0.005, 34, 162, 1.5, 2.5, 4.5)$arl,
  arl1 = dsnp_arl(0.0075, 34, 162, 1.5, 2.5, 4.5)$arl,
  ass0 = dsnp_ass(0.005, 34, 162, 1.5, 2.5)$ass
)
```

# Generalized variance chart

Let \(S\) be the sample covariance matrix from a p-variate normal subgroup of
size \(n\). Then

\[
(n-1)S\sim W_p(n-1,\Sigma).
\]

The monitored statistic is

\[
G=|S|.
\]

Bartlett's decomposition gives

\[
G\overset{d}{=}
\frac{|\Sigma|}{(n-1)^p}
\prod_{j=1}^{p}Y_j,
\qquad
Y_j\sim\chi^2_{n-j},
\]

with independent factors.

```{r gv-example}
gv_table <- do.call(
  rbind,
  lapply(c("normal", "cf", "exact"), function(method) {
    lim <- gv_limits(
      n = 10,
      p = 2,
      det_sigma = 0.5320,
      type = method
    )
    risk <- gv_alpha_risk(
      n = 10,
      p = 2,
      det_sigma = 0.5320,
      type = method
    )
    data.frame(
      method = method,
      lcl = lim$lcl,
      center = lim$center,
      ucl = lim$ucl,
      alpha = risk$alpha,
      arl0 = risk$arl0
    )
  })
)
gv_table
```

```{r gv-figure}
set.seed(321)
x <- array(rnorm(10 * 8 * 2), dim = c(10, 8, 2))
cchart.GV(x, Sigma = diag(2), type = "exact", plot = TRUE)
```

For dimension two, the exact distribution simplifies to

\[
G\overset{d}{=}
\frac{|\Sigma|}{4(n-1)^2}
\left(\chi^2_{2n-4}\right)^2.
\]

For higher dimensions, IQCC offers simulation using the Bartlett factors. The
package does not claim a generic Meijer-G implementation. Published exact
quantiles are used only in explicitly supported dimension-three cases.

# Auxiliary trace chart

The generalized variance chart is sensitive to determinant changes in the
covariance matrix. A complementary diagnostic monitors the trace of the
standardized covariance matrix,

\[
T=(n-1)\operatorname{tr}(\Sigma_0^{-1}S).
\]

If the process is multivariate normal and \(\Sigma=\Sigma_0\), then

\[
T\sim\chi^2_{p(n-1)}.
\]

This gives an exact upper control limit for the trace chart:

```{r trv-limits}
trv_limits(n = 8, p = 2, alpha = 0.0027)
```

The trace chart complements `cchart.GV()` because covariance changes can
preserve the determinant while changing the sum of standardized eigenvalues.
The following deterministic example has two subgroup covariance matrices with
the same determinant; `|S|` is unchanged, while `tr(V)` increases.

```{r trv-same-determinant}
base <- sqrt(3 / 4) * rbind(
  c(1, 1), c(1, -1), c(-1, 1), c(-1, -1)
)
g0 <- base %*% chol(diag(2))
g1 <- base %*% chol(diag(c(4, 0.25)))

data.frame(
  generalized_variance = gv_stat(list(g0, g1)),
  trace_statistic = trv_stat(list(g0, g1), Sigma0 = diag(2))
)
```

```{r execution-marker, results='asis'}
cat("<!-- IQCC_EXECUTED_FOUNDATIONS -->\n")
```

# Validation strategy

IQCC combines algebraic property tests, independently coded numerical oracles,
and reproduction of published numerical examples. This is particularly
important because a plausible formula may still use a different tail
convention, integer threshold, or parameter-estimation rule than the intended
method.

# References

- Joekes, S. and Barbosa, E. P. (2013). An improved attribute control chart for monitoring non-conforming proportion in high quality processes. *Control Engineering Practice*, 21, 407-412.
- Joekes, S., Smrekar, M. and Barbosa, E. P. (2015). Extending a double sampling control chart for non-conforming proportion in high quality processes to the case of small samples. *Statistical Methodology*, 23, 35-49.
- Barbosa, E. P., Gneri, M. A. and Meneguetti, A. *Improving Shewhart-type Generalized Variance Control Charts for Multivariate Process Variability Monitoring using Cornish-Fisher Quantile Correction, Meijer-G Function and Other Tools*.
