---
title: "Monitoring process variability with exact limits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monitoring process variability with exact limits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Purpose

Process monitoring often starts with the mean, but changes in variability can be
just as important. A process may keep approximately the same center while its
spread increases, decreases, or changes structure. In such cases, a variability
chart can detect deterioration that an X-bar chart may miss.

IQCC includes improved tools for univariate variability monitoring. The package
focuses on cases where familiar Shewhart-type limits are easy to draw but not
well calibrated for the actual sampling distribution of the statistic.

## Why exact limits matter

Range and standard deviation charts are based on statistics with asymmetric
sampling distributions, especially for small subgroups. Normal-style symmetric
limits can therefore produce a false alarm probability different from the
nominal value. IQCC addresses this by providing probability limits based on the
sampling distribution of the statistic whenever possible.

The two central univariate variability charts are:

- `cchart.R()` for range monitoring;
- `cchart.S()` for standard deviation monitoring.

## Range charts

The classical R chart monitors subgroup ranges. IQCC supports the ordinary
Shewhart-style R chart and an exact R chart based on the distribution of the
relative range. The exact chart is requested with `type = "tukey"`.

```{r r-chart-example, eval = FALSE}
library(IQCC)

data(pistonrings)

# Classical R chart for Phase I data
cchart.R(pistonrings[1:25, ], n = 5)

# Exact R chart for prospective data, using Phase I data to estimate sigma
cchart.R(
  x = pistonrings[26:40, ],
  n = 5,
  type = "tukey",
  y = pistonrings[1:25, ]
)
```

A useful diagnostic companion is `alpha.risk()`, which computes the actual false
alarm probability for the classical three-sigma R chart for a given subgroup
size.

```{r alpha-risk-example}
library(IQCC)

alpha.risk(5)
```

This is useful when deciding whether the ordinary R chart is sufficiently well
calibrated for a particular subgroup size.

## S charts

The S chart monitors subgroup standard deviations. IQCC supports a normalized S
chart and an exact S chart. The exact limits use the relationship between the
sample variance and the chi-square distribution.

```{r s-chart-example, eval = FALSE}
library(IQCC)

data(softdrink)

# Normalized S chart
cchart.S(softdrink, type = "n")

# Exact S chart; m is the subgroup size
cchart.S(softdrink, type = "e", m = 10)
```

Use the exact option when subgroup sizes are small enough that the asymmetry of
the sampling distribution matters for the control limits.

## Choosing between R and S charts

The R chart is simple and historically common. It is often attractive for small
subgroups because the range is easy to compute and explain. The S chart uses
more information from the subgroup and is generally more directly connected to
variance estimation.

A practical rule is:

- use `cchart.R(type = "tukey")` when the workflow is range-based and the goal is
  to keep the R-chart interpretation while improving limit calibration;
- use `cchart.S(type = "e")` when subgroup standard deviations are the natural
  statistic and exact probability limits are desired.

## Relation to multivariate variability

Univariate variability monitoring tracks the spread of one quality
characteristic. Multivariate variability monitoring is harder because changes may
occur in individual variances, covariances, or both.

The current development roadmap includes generalized variance charts based on
`|S|` and auxiliary trace charts based on `tr(V)`. These future methods are meant
to extend the same principle used by IQCC's univariate variability charts: prefer
probability-calibrated limits when the usual approximations are not reliable.

## Practical workflow

A typical IQCC workflow for variability monitoring is:

1. Start with the standard chart that matches the statistic used in the process
   history: R for ranges, S for subgroup standard deviations.
2. Use the exact IQCC option when the subgroup size is small or the false alarm
   risk of the classical chart is a concern.
3. Use `alpha.risk()` to quantify how different the classical R-chart false
   alarm probability is from the nominal value.
4. For multivariate variability, treat the current IQCC implementation as a
   roadmap: generalized variance and trace-based charts are planned extensions,
   not yet part of the stable user-facing API.

## Summary

IQCC is most useful for variability monitoring when the question is not merely
"can I draw an R or S chart?", but rather "are the limits on this chart
calibrated for the actual distribution and subgroup size?" The exact R and S
chart options provide a more statistically faithful alternative while preserving
the familiar control-chart workflow.
